---
title: "College Data"
output: html_document
date: "2024-05-15"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(xgboost)
library(caret)
library(cvms)
library(caTools)
library(ggcorrplot)
library(randomForest)
```

## R Markdown


```{r}

df<-read.csv("college.csv")

```




```{r}
#df$W.s[df$W.s > 0] <- 1 
df$Pd<-df$Tm./df$Opp.

dfyears<-df

cols=c("W.L.","SRS","SOS","Pd","FG.","X3P.","X3P","FT","FT.","ORB","TRB","AST","STL","BLK","TOV","W.s")


df<-select(df,cols)

cols=c("W.L.","SRS","SOS","Pd","FG.","X3P.","X3P","FT","FT.","ORB","TRB","AST","STL","BLK","TOV","W.s","year","Team")

dfyears<-select(dfyears,cols)

df2023<-filter(dfyears,year==2023)
cols=c("W.L.","SRS","SOS","Pd","FG.","X3P.","X3P","FT","FT.","ORB","TRB","AST","STL","BLK","TOV","W.s")
df2023<-select(df2023,cols)
```
```{r}

sum(!complete.cases(df))


sample<- sample.int(n=nrow(df),size=floor(.8*nrow(df)),replace=FALSE)
train<-df[sample,]
test<-df[-sample,]
testteams<-dfyears[-sample,]


x_test<-test[,-16]
y_test<-test[,16]
x_train<-train[,-16]
y_train<-train[,16]
```
```{r}
corr <- round(cor(df), 1)
ggcorrplot(corr)
```
lets use some differnce AI's

```{r}
#go back to number wins 
model <- randomForest(
  formula = y_train ~ .,
  data = x_train
)
model
which.min(model$mse)
sqrt(model$mse[which.min(model$mse)]) 
plot(model)
```

```{r}
varImpPlot(model) 
model_tuned <- tuneRF(
               x=x_train[,-16], #define predictor variables
               y=y_train, #define response variable
               ntreeTry=500,
               mtryStart=4, 
               stepFactor=1.5,
               improve=0.01,
               trace=FALSE #don't show real-time progress
               )


```


```{r}

predictions<-data.frame(predict(model, newdata=x_test))
predictions$team<-testteams$Team
predictions$win<-predictions[[1]]
#predictions<-filter(predictions,win>=1)

#winners<-select(filter(teams,W.s>0),Team)

predictions$win[predictions$win < 1.1] <- 0
predictions$win[predictions$win >= 1.1] <- 1 


df2023$W.s[df2023$W.s > 0] <- 1 
y_test[y_test>0]<-1




#if we only select predicted >1

#predictions<-filter(predictions,win>=1)
#y_test<-y_test[y_test>0]

#select the same 20 teams from df2023 that are still in preditions


predictions$win
y_test
accuracy <- sum(predictions$win == y_test) / nrow(predictions)
accuracy
```
test over more years to gain sample size confidence
```{r}
predictions<-data.frame(predict(model, newdata=x_test))
predictions$team<-testteams$Team
predictions$win<-predictions[[1]]
#predictions<-filter(predictions,win>=1)

#winners<-select(filter(teams,W.s>0),Team)

predictions$win[predictions$win < 1.1] <- 0
predictions$win[predictions$win >= 1.1] <- 1 

```

100% accuracy for 18 of 30 games loaded in, useing regression predicter greater than 1.1 games won in march madness using random forest



```{r}
confusionMatrix(as.factor(predictions$win), factor(df2023$W.s,levels=0:6))
```



```{r}

cm <- confusionMatrix(factor(predictions$win), factor(df2023$W.s,levels = 0:6))
cfm <- as_tibble(cm$table)
plot_confusion_matrix(cfm, target_col = "Reference", prediction_col = "Prediction", counts_col = "n")

```
```{r}
dfWinners<-filter(teams,W.s>0)

D2 <- predictions %>% dplyr::filter(!team %in% dfWinners$Team)

D2
```
The one miss, was a team the BARLEY made the "1" mark to be considered a good pick


```{r}



xgb_train<-xgb.DMatrix(data=as.matrix(x_train),label=y_train)
xgb_test<-xgb.DMatrix(data=as.matrix(x_test),label=y_test)

xgb_params <- list(
  booster="gbtree",
  eta=0.01,
  max_depth=8,
  gamma=4,
  colsample_bytree=1,
  objective="multi:softprob",
  eval_metric="mlogloss",
  num_class=length(levels(as.factor(df$W.s)))) 

```


```{r}
xgb_model<-xgb.train(params=xgb_params,data=xgb_train,nround=5000,verbose=1)



```




start with srs,sos,mov,hob,ft/fga,
add more data

add code from desktop to jup asap



```{r}
feature_scores<-xgb.importance(feature_names = colnames(xgb_train),model=xgb_model)
feature_scores
```



wow, really need more features


if you want to plot feature scores...
```{r}

xgb.plot.importance(feature_scores)
```


onto predictions


```{r}
x_test<-df2023[,-16]
xgb_preds <- predict(xgb_model, as.matrix(x_test), reshape = TRUE)#reminder, to say, predict the upcoming march, put that seasons data in x_test HERE
xgb_preds <- as.data.frame(xgb_preds)
colnames(xgb_preds) <- levels(factor(df$W.s))
#xgb_preds

predicter<-xgb_preds
teams<-filter(dfyears,year==2023)
predicter$team<-teams$Team

xgb_preds$PredictedClass <- apply(xgb_preds, 1, function(y) colnames(xgb_preds)[which.max(y)])

predicter
#UNCOMMENT THIS CODE BELOW WpreiHEN TESTING OLD DATA, AND NOT PREDICTING

#xgb_preds$ActualClass <- levels(factor(df$W.s))[y_test + 1]
#xgb_preds

predicter$Win<-predicter[[2]]
predicter$Team<-predicter$team
predicter<-filter(predicter,Win>=.75)
predicter
```

a 75% confidence level successfully predicted the outcome of 14/15 march madness round 1 games in 2023



```{r}
accuracy <- sum(xgb_preds$PredictedClass == xgb_preds$ActualClass) / nrow(xgb_preds)
accuracy

```

check the confuse matrix
```{r}
confusionMatrix(as.factor(xgb_preds$ActualClass), factor(xgb_preds$PredictedClass,levels=0:6))
```
```{r}

cm <- confusionMatrix(factor(xgb_preds$ActualClass), factor(xgb_preds$PredictedClass,levels = 0:6))
cfm <- as_tibble(cm$table)
plot_confusion_matrix(cfm, target_col = "Reference", prediction_col = "Prediction", counts_col = "n")

```


Dont mind the roullete sim down here, had to show a friend their strategy sucked
```{r }
#3 outcomes can happen 

#we make four bets, call them bet_col1 


```

```{r}
get_rolls <- function()
{
  set.seed(NULL)
  return <- sample(1:38,1000,replace=TRUE)
}


```

```{r}
get_bets <- function(ROLL,bet1,bets1,bet2,bets2,bet3,bets3,bet4,bets4)
{
for (i in 1:1000)
{
  if (ROLL[i] %in% bet1) #bet hit
    {
      bets1[i]<-TRUE
  }
  else
  {
      bets1[i]<-FALSE
  }
   if (ROLL[i] %in% bet2) #bet hit
    {
      bets2[i]<-TRUE
  }
  else
  {
      bets2[i]<-FALSE
  }
   if (ROLL[i] %in% bet3) #bet hit
    {
      bets3[i]<-TRUE
  }
  else
  {
      bets3[i]<-FALSE
  }
   if (ROLL[i] %in% bet4) #bet hit
    {
      bets4[i]<-TRUE
  }
  else
  {
      bets4[i]<-FALSE
  }
}
  return <- list(bets1,bets2,bets3,bets4)
}


```


```{r}
big_bets <- function(bets1,bets2,bets3,bets4)
{
bigbets<-c()

for (i in 1:1000)
{
  if((((bets1[i]==TRUE )| (bets2[i])==TRUE)  )&  (((bets3[i]==TRUE) | (bets4[i])==TRUE)  ) )
  {
    bigbets[i]<-TRUE
  }
  else
  {
    bigbets[i]<-FALSE
  }
}
return <-bigbets 
}
```


```{r}
small_bets <-function(bets1,bets2,bets3,bets4)
{
smallbets<-c()

for (i in 1:1000)
{
  if(  (bets1[i]==TRUE) | (bets2[i]==TRUE)|(bets3[i]==TRUE) | (bets4[i]==TRUE)  )
  {
    smallbets[i]<-TRUE
  }
  else
  {
    smallbets[i]<-FALSE
  }
}
return <- smallbets
}
```


```{r}
run_sim <-function(bigbets,smallbets)
{
start<-100
amount<-100
final<-start
results<-c()
x<-0
up<-FALSE
for (i in 1:1000)
{
  if(final<start)
  {
    amount<-start-final
  }
  
  if(bigbets[i] == TRUE)
  {
    final<-(final+(1.5*amount))
  }
  else if (smallbets[i]==TRUE)
  {
    final<-(final- (.25*amount))
  }
  else
  {
    final <- final-(amount)
  }
  if(final <= 99)
  {
    x<-x+1
    #print("you lost all your money after {i} hands")
    results[x]<-(sprintf("You can no longer bet your amount after %s Rolls\n",i))
    break
  }
  if(final > start)
  {
    x<-x+1
    profit<-final-start
    results[x]<-(sprintf("you currently have more than you started by %s \n",profit))
    
    up<-TRUE
    break
  }

  #print(paste(sprintf("After %s rolls you have",i), sprintf("$%s dollars left",final), sep = " "))
}
message(results)
return <- (final-start)

}

```

## Including Plots

You can also embed plots, for example:

```{r}
total<-0
values<-seq(1,38,1)

bet1<-seq(1,12,1)
bet2<-seq(13,24,1)

bet3<-c(1,34,3)
bet4<-c(2,35,3)
ups=c()
results=c()
for (i in 1:100)
{
ROLLS <- get_rolls()
bets1<-c()
bets2<-c()
bets3<-c()
bets4<-c()
bets<-get_bets(ROLLS,bet1,bets1,bet2,bets2,bet3,bets3,bet4,bets4)
bigbets<-big_bets(bets[[1]],bets[[2]],bets[[3]],bets[[4]])
smallbets<-small_bets(bets[[1]],bets[[2]],bets[[3]],bets[[4]])
ups[i]<-run_sim(bigbets,smallbets)
Sys.sleep(.01)
}
print(mean(ups))
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
